Automatic discovery of cell types and microcircuitry from neural connectomics

### 2014-06-25 docs/paper.tex
updates to all

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses, in a
  principled and probabilstically-coherent manner. We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  C. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{An overview of our method. a. Input data is a connectivity
    matrix indicating whether cell $i$ (row) synapses on cell $j$
    (column), a distance matrix between cell soma, b. the depth of the
    cell soma, and the c. depth profile of the cells synaptic
    connectivity. d. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. d. shows the cell connectivity matrix with cells of the
    same type grouped together. e. shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. f. shows the probability of connection between two cell types
that very rarely connect -- there's a background ``base'' connection rate
to account for errors in data, but the probability is very low. 
We can plot this these types g.) to show the relationship and spatial
connections, and how probable various types are to connect to one
another. h. The connectivity between our discovered types changes as a
function of distance. }

@@

From these assumptions (priors) develop a generative Bayesian model
that estimates the underlying cell types and how they connect. We take
as input (fig ~\ref{fig:overview}a) the connectivity matrix of cells,
a matrix of the distance between cells, and the per-cell soma depth
(fig ~\ref{fig:overview}b) and the depth profile of the cell's
synapses (fig ~\ref{fig:overview}c). We perform joint probabilistic
inference to automatically learn the number of cell types, which cells
belong to which type, their type-specific connectivity, and how
connections between types vary with distance. We also simultaneously
learn the soma depth associated with each type and the typical
synaptic density profile (fig ~\ref{fig:overview}d.).

@@

To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and comparing the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  Indeed, it does far better than the
existing infinite stochastic block model that assumes that there is no
distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). Thus we can apply our model to simulated datasets with structure
and scale similar to that of our biological datasets and recover the known 
correct structure.

@@

\subsubsection{Mouse Retina}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{a. Cell-cell assignment matrix reflecting the posterior
    distribution of clusterings found by our model. Large contiguous
    black areas are cells that were considered the same type in the
    majority of samples. b. The cells and their true types sorted to
    agree with the coassignment matrix. c. A comparison of the
    predictive accuracy (Area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information. d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in C. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{C. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{C. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). d.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. e.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:otherdata}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by Brenner in
1986.

@@

\subsection{Microprocessor}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:otherdata}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:otherdata}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

\section{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered from connectomics data.  We have
shown its applicability to a newly published connectomics dataset from
the mouse retina. When compared to existing methods, we show how the
incorporation of all of this data yields results that combine both
high link-prediction accuracy and high agreement with human
anatomists.  I also reveals interesting structure when applied to the
classical connectivity of the worm c. elegans, and the connectivity of
a historical man-made microprocessor. We have found that combining the
available data types allows us to discover cell types and
microcircuitry that were known to exist in the systems based on
decades of previous research and allows good prediction of
connectivity.

@@

Eventually, algorithms will have to become a central tool for
anatomists, as it will progressively become impossible for humans to
parse the huge datasets. This transition may follow a similar
transition to the one underway in molecular biology (with gene-finding
algorithms) and evolutionary biology with (computational
phylogenetics).  In fact, it is known that there are significant
disagreements across human anatomists and computational approaches
promise to make the debates more data driven.

@@

Inference is performed in three steps via composible transition 
kernels -- one for structural, one for per-type parameters, and
one kernel for global parameters and hyperparameters. Details
of data preprocessing, inference parameters, and runtime can
be found in the Methods section.

@@

 \item \textbf{Acknowledgments} We thank Josh Vogelstein for discussions and reading of the manuscript, Finale Doshi-Velez for early discussions on the model, and Erica Peterson, and Jonathan Glidden for extensive manuscript review. Funding for compute time was provided by Amazon Web Services ``AWS in Education'' grants. 
\item \textbf{Author Contributions} KK and EJ developed model. EJ derived inference, implemented code, tested, and ran experiments. KK and EJ wrote manuscript text and solicited feedback. 
 \item \textbf{Competing Interests} The authors declare that they have no
competing financial interests.
 \item \textbf{Correspondence} Correspondence and requests for materials
should be addressed to E.J.~(email: jonas@eecs.berkeley.edu).
\end{itemize}

@@

We chain these three kernels together, and then globally anneal on the
likelihood from a temperature of $T=64$ down to $T=1$ over $900$
iterations unless otherwise indicated, and then run the chain for
another $100$ iterations. We then generate at least $20$ samples, each
taken from the end of a single Markov chain initialized from different
random initial points in the state space. For visualization we pick the
chain with the highest log likelihood, but for all numerical
comparisons (including link probability and ARI) we use this full
collection of samples from the posterior distribution to estimate the
resulting statistics.

@@

\subsection{Model Comparison}
We compare our model with a standard network clustering model, the
latent-position clutsering model. This model assumes each cell belongs
to one of K clusters, and each cluster is associated with a
$d-$dimensional Gaussian distribution. The probability of a link is
then a function of the distance between the data points in this
continuous-space. We use \autocite{VBLPCM} a variational
implementation provided in R, parametrically varying the number of
latent dimensions and the number of requested groups.  While this
model provides reasonable link predictive accuracy, the clusterings
dramatically disagree with those from human anatomists.

@@

Figures \ref{fig:mixing:ari} and \ref{fig:mixing:scorevstime} shows
the ARI to ground truth and the total log score as a function of
runtime.  We see dramatic changes in log score initially as we vary
the temperature, stabilizing as runtime progresses, for each
chain. Then we see the characteristic jumps between nearby modes
towards the end of the run, in both log score and ARI.  Importantly,
regardless of whether our model over- or under-estimates the exact
posterior variance about the network, we find points in the latent
variable space that are both predictive \textit{and} parsimonious,
largely agreeing with the human anatomists and predicting existing
connections.

>>>>

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses, in a
  principled and probabilistically-coherent manner. We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  C. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{Deriving circuitry and types from connectomics data a. As input we take the connectivity betwen cells (b), the distance between them (c), the depth of the cell bodies (d), and the depth profile of the syanpses (e). f. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell types
that very rarely connect -- there's a background ``base'' connection rate
to account for errors in data, but the probability is very low. 
We can plot this these types (i) to show the relationship and spatial
connections, and how probable various types are to connect to one
another. j.) The connectivity between our discovered types changes as a
function of distance. }

@@

From these assumptions (priors) we develop a generative Bayesian model
that estimates the underlying cell types and how they connect. We take
as input (fig ~\ref{fig:overview}a) the connectivity matrix of cells (fig ~\ref{fig:overview}b) ,
a matrix of the distance between cells (fig ~\ref{fig:overview}c), the per-cell soma depth
(fig ~\ref{fig:overview}d) and the depth profile of the cell's
synapses (fig ~\ref{fig:overview}e). We perform joint probabilistic
inference to automatically learn the number of cell types, which cells
belong to which type, their type-specific connectivity, and how
connections between types vary with distance. We also simultaneously
learn the soma depth associated with each type and the typical
synaptic density profile (fig ~\ref{fig:overview}f-h.).

@@

To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and comparing the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, existing infinite stochastic block model assumes cell type alone matters, and thus finds small neighborhoods of connected nodes (instead of global connectivity patterns). The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). Thus we can apply our model to simulated datasets with structure
and scale similar to that of our biological datasets and recover the known 
correct structure.

@@

\subsubsection{Learning types and circuitry in the retina}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{\textbf{Visualizing type inference uncertainty.} Our fully Bayesian model gives a confidence estimate (posterior probability) that any two given cells are of the same type. In b.) we visualize that cell-cell coassignment matrix, showing the probability that cell i is of the same type as cell j on a range from 0.0 to 1.0. The block structure shows subsets of cells which are believed to all belong to the same type. For comparison, a.) shows the anatomist-defined type for each cell, grouped broadly into the coarse types identified in the previous figure. \textbf{Link vs cluster accuracy} c. A comparison of the
    predictive accuracy (area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information, as well as the clustering accuracy. Note that our model sacrifies very little predictive accuracy for additional clustering accuracy. By comparison, conventional methods fail at one or both.  d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}

@@

\subsection{Recovering spatial connectivity in multiple graphs simultaneously}

Having shown our model to work on the repeating tessellated, laminar structure of the mammilian retina, we then apply our model to a structurally very different connectome -- the whole body of a small roundworm: 
\textit{Caenorhabditis elegans} is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in C. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia. C. elegans also differs from the retina in that the measured connectome is actually two separate graphs -- one of directed chemical synapses and another of undirected electrical synapses.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=4.5in]{celegans.ai}}
  \caption{Discovering connectivity and type in \textit{C. elegans}. 
 a.) Initial
    \textit{C. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). }
  \label{fig:celegans}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:celegans}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by White in
1986.

@@

\subsection{Types and connectivity in artificial structures}
\begin{figure}
  \centering 
  \centerline{\includegraphics[width=4in]{mos6502.ai}}
  \caption{Discovering connectivity and type the 6502 microprocessor.
    a.) is the micrograph of the original microprocessor, with the
    region containing the registers under study highlighted. b.) Our
    graph consists of the interconnections of MOS field-effect
    transistors with three terminals, Gate, C1, and C2. Reconstruction
    technique did not permit resolution of C1 and C2 into source and
    drain. c.) The spatial distribution of the transistors in each
    cluster show a clear pattern d.) The clusters and connectivity
    versus distance for connections between Gate and C1, Gate and C2,
    and C1 and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus (SDB).}
  \label{fig:mos6502}
\end{figure}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:mos6502}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:mos6502}c, showing the man-made horizontal and
vertical layout of the same types of transistors.

\section{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered from connectomics data.  We have
shown its applicability to regularly structured laminar neural circuits like the retina, 
as well as a less structured whole neuronal organism (\textit{C. elegans}) and an artificial processor. When compared to existing methods, we show how the
incorporation of all of this data yields results that combine both
high link-prediction accuracy and high agreement with human
anatomists. We have found that combining the
available data types allows us to discover cell types and
microcircuitry that were known to exist in the systems based on
decades of previous research and allows good prediction of
connectivity.

@@

Our work shows how rich probabilistic models can contribute to computational neuroanatomy. 
Eventually, algorithms will have to become a central tool for
anatomists, as it will progressively become impossible for humans to
parse the huge datasets. This transition may follow a similar
transition to that of molecular biology (with gene-finding
algorithms) and evolutionary biology with (computational
phylogenetics). Ultimately, computational approaches may help resolve the significant
disagreements across human anatomists.

@@

Inference is performed in three steps via composable transition 
kernels -- one for structural, one for per-type parameters, and
one kernel for global parameters and hyperparameters. Details
of data preprocessing, inference parameters, and runtime can
be found in the Methods section.

@@

 \item \textbf{Acknowledgments} We thank Josh Vogelstein for discussions and reading of the manuscript, Finale Doshi-Velez for early discussions on the model, and Erica Peterson, and Jonathan Glidden, and Yarden Katz for extensive manuscript review. Funding for compute time was provided by Amazon Web Services ``AWS in Education'' grants. 
\item \textbf{Author Contributions} KK and EJ developed model. EJ derived inference, implemented code, tested, and ran experiments. KK and EJ wrote manuscript text and solicited feedback. 
 \item \textbf{Competing Interests} The authors declare that they have no
competing financial interests.
 \item \textbf{Correspondence} Correspondence and requests for materials
should be addressed to E.J.~(email: jonas@eecs.berkeley.edu).
\end{itemize}

@@

We chain these three kernels together, and then globally anneal on the
likelihood from a temperature of $T=64$ down to $T=1$ over $900$
iterations unless otherwise indicated, and then run the chain for
another $100$ iterations. We then generate at least $20$ samples, each
taken from the end of a single Markov chain initialized from different
random initial points in the state space. For visualization we pick the
chain with the highest log likelihood, but for all numerical
comparisons (including link probability and cluster accuracy) we use this full
collection of samples from the posterior distribution to estimate the
resulting statistics.

@@

\subsection{Model Comparison}
We compare our model with a standard network clustering model, the
latent-position clustering model. This model assumes each cell belongs
to one of K clusters, and each cluster is associated with a
$d-$dimensional Gaussian distribution. The probability of a link is
then a function of the distance between the data points in this
continuous-space. We use \autocite{Salter-Townshend2013} a variational
implementation provided in R, parametrically varying the number of
latent dimensions and the number of requested groups.  While this
model provides reasonable link predictive accuracy, the clusterings
dramatically disagree with those from human anatomists.

@@

Figures \ref{fig:mixing:ari} and \ref{fig:mixing:scorevstime} shows
the cluster accuracy (ARI) to ground truth and the total log score as a function of
runtime.  We see dramatic changes in log score initially as we vary
the temperature, stabilizing as runtime progresses, for each
chain. Then we see the characteristic jumps between nearby modes
towards the end of the run, in both log score and ARI.  Importantly,
regardless of whether our model over- or under-estimates the exact
posterior variance about the network, we find points in the latent
variable space that are both predictive \textit{and} parsimonious,
largely agreeing with the human anatomists and predicting existing
connections.
